Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-31 manuscript.rmd
Edited symbols

\
Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear. While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high Kmax) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

\
Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

@@

where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.

@@

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib; daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (*n* = 5, cylinder volume = 1250 cm{3}). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \ref(tab:param) for best-fit parameter ranges. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (g cm^{-3}^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (g m^{-2}^) for leaves receiving direct sunlight, respectively.  Fitted polynomial equations are described as,

@@

\
\begin{equation}
K_{leaf,z,n} = Ae^{-B\Psi_{z,n}}
(\#eq:k-soil)
\end{equation}

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diamater at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^{-1})) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census onterval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

@@

For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive $\rho$ and *R*^2^ >= 0.1.

@@

\
The model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (\@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See [Table S1]{erd-sp} for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, Stem}$), leaf Turgor Loss Point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{x, ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max, Stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf Turgor Loss Point ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island. See [Table S1]{erd-sp} for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: Kmax units as in Fig. 2. Water Potentials are in Mpa.]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{x,~ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{x,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` [`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`] (median [95% CI]) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

```{r erd-mort-changed, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time-new, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{x, z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' strategy axis by demonstrating a link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes. On the aquisitive end were species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable, abundant source of water at depth, which actualised to a lower associated mortality-risk, whilst at the slow end were species with inefficient, but safer xylem, associated with a water environment of high variability.

@@

fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high Kmax) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;]. Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012]. Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~stem}$) in two angiosperm systems. High Kmax is associated with photosynthetic rates with faster growth potential.

>>>>

\
Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear. While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

\
Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specifically, (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

@@

where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), $\theta_{s}$ is the saturated volumetric water content (0.51, cm^3^cm^-3^), and $b$, an index for soil pore-size distribution (10, unitless), respectively.

@@

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib); daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \@ref(tab:param) for best-fit parameter ranges. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight, respectively.  Fitted polynomial equations are described as,

@@

\
\begin{equation}
K_{leaf,z,n} = Ae^{-B\Psi_{soil,~z,n}}
(\#eq:k-soil)
\end{equation}

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diamater at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^-1^)) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census onterval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardised (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

@@

For each model \@ref(eq:M1)- \@ref(eq:M4) and species combination ($sp$), we obtained goodness-of-fit (*R*^2^) and Pearson's $\rho$ for all soil depths $z$, and for each sp, model combination we chose the $z$ with highest *R*^2^ as the Effective Rooting Depth (ERD); after retaining species with positive Pearson's $\rho$ and *R*^2^ >= 0.1.

@@

\
The model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2)) had highest goodness-of fit (*R*^2^ = 0.51; *p* = 0.072), followed by followed by model without both VPD and and Leaf Fraction (Eq. \@ref(eq:M1); Fig. S\@ref(fig:other-erd-mods)). Models with leaf fraction had only three species with sufficient species-level goodness-of-fit (Eqs. \@ref(eq:M3)-\@ref(eq:M4)) versus seven for models without (Eqs. \@ref(eq:M1)-\@ref(eq:M2)). For rest of the analyses we therefore used species ERD estimates from the model with VPD and without Leaf Fraction (Eq. \@ref(eq:M2); Fig. \@ref(fig:erd-iso)).

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

To evaluate whether ERD was associated with above-ground hydraulic traits (*n* = 8 species with both datasets), we regressed species ERD with maximum stem hydraulic conductance ($K_{max, stem}$), leaf turgor loss point ($\Psi_{tlp}$), vulnerability to embolism from cavitation measured in terms of pressure at which 88% of stem hydraulic conductivity is lost ($\Psi_{88,~stem}$), and hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$).

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

\
In this forest modelled Effective Rooting Depths (ERD) for the `r length(erd.sp)` large, canopy tree species varied from 0.5 m to 13 m (Figure \@ref(fig:sp-erd)). For the subset of species for which hydraulic properties were available, species with deeper ERD showed greater maximum stem hydraulic conductance ($K_{max,~stem}$; *R*^2^ = 0.39, *P* = 0.098, *n* = 8; Figure \@ref(fig:erd-hyd)A), less negative leaf  ($\Psi_{tlp}$); *R*^2^ = 0.6, p = 0.024, *n* = 8; Figure \@ref(fig:erd-hyd)B), greater vulnerability to embolism from cavitation with less negative ($\Psi_{88,~stem}$) (*R*^2^ = 0.72, p = 0.007, *n* = 8; Figure \@ref(fig:erd-hyd)C), and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$; *R*^2^ = 0.74, p = 0.013, *n* = 8; Figure \@ref(fig:erd-hyd)D).

```{r sp-erd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap= paste0("\\label{lab-sp-erd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.sp)," large, canopy species of Barro Colorado Island. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/ERD_by_sp_large_canopy.jpeg"))
```

```{r erd-hyd, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "60%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-hyd}Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus hydraulic propertiese for eight large, canopy species found on Barro Colorado Island (Panama). *R*^2^ and *P*-values are given in panel insets.[X-Axis labels to be added: $K_{max}$ units as in Fig. 2. Water Potentials are in MPa.]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/Select_traits_depth.jpeg"))
```

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

```{r erd-mort, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-mort}Mortality rate (mean $\\pm$ 1SE) versus Modelled Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) for ", length(erd.mrate.sp)," canopy species found on Barro Colorado Island. *R*^2^ and *P*-values are given in panel insets.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mortality_rate_by rdi.gr.jpeg"))
```

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time} Modelled Species Effective Rooting Depth (ERD; mean $\\pm$ (add error) in m) versus time spent below critical hydraulic threhold--the number of days for which $\\Psi_{soil,~z}$ was below $\\Psi_{crit}$, defined as $\\Psi_{80,~leaf}$, over 1990-2015, where z is the soil depth matching species ERD--for ", length(erd.mrate.sp), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are jittered vertically around the named depths (ticks) to be able to distinguish visually. (Are *R*^2^ and *P*-values needed here?)")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were species with hydraulically efficient, but risky suit of hydraulic traits. These were associated with, and likely optimised for, an environment of reliable and abundant source of water at depth, which actualised to a lower associated mortality-risk. At the slow end were species with inefficient, but safer xylem, associated with a shallow soil- water environment of high variability.

@@

fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high $K_{max}$) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;]. Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012]. Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~stem}$) in two angiosperm systems. High $K_{max}$ is associated with photosynthetic rates with faster growth potential.
